Categories of Optics

### 2018-03-02 paper.tex
Fix some old references to prethings

\begin{remark}
In this section we will focus only on preoptics of the form $p : (S,S) \hto (A, A)$. We therefore abbreviate $\Optic_\M((S, S), (A, A))$ as $\Optic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.
\end{remark}

For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $outside : \Optic_\M(S, A) \to \C(S, S)$. By composition in the other direction we have a map $\C(S, M A) \times \C(M A, S) \to \C(M A, M A)$. Including this object $\C(M A, M A)$ into the coend $\int^{M \in \M} \C(M A, M A)$, we induce a map $inside : \Optic_\M(S, A) \to \int^{M \in \M} \C(M A, M A)$. 
\begin{definition}
An preoptic $p : S \hto A$ is an \emph{optic} if $outside(p) = \id_S$, and $inside(p) = \langle t A \rangle$ for some $t : M \to M$ in $\M$.
\end{definition}

\begin{proposition}
There is an subcategory $\Lawful_\C$ of $\Optic_\C$ given by objects of the form $(X, X)$ and lawful optics between them.
\end{proposition}
\begin{proof}
The identity preoptic $\iota(\id_X, \id_X)$ is clearly an optic.

Suppose we have two optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$, and such that $lr = \phi S$ and $l'r' = \psi A$. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is also an optic:
\begin{align*}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\intertext{and}
(Ml')l r (Mr') &= (Ml')(\phi S)(Mr') \\ 
&= (\phi NA) M(l')M(r') \\ 
&= (\phi NA) M(l'r') \\
&= (\phi NA) M(\psi A) \\
&= \phi(\psi A)
\end{align*}
by the naturality of the action of $\M$.
\end{proof}

@@

In view of this Proposition, whenever we posit a representative $\langle l, r \rangle$ of an optic, we may assume that $lr = \phi A$ on the nose, for some $\phi : M \to M$.

@@

\begin{definition}
Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ is the category of optics with respect to this action: $\Lens = \Lawful_\times$.
\end{definition}

As remarked in the introduction \todo{(todo: remark in the introduction)}, the preoptics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\Optic_\times(S, A) &= \int^{M \in \M} \C(S, M \times A) \times \C(M \times A, S) \\
&\cong \int^{M \in \M} \C(S, M) \times \C(S, A) \times \C(M \times A, S) && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A, S) && \text{(co-Yoneda)}
\end{align*}

@@

This has been written out many times before in the case when $lr$ is the identity \cite{IsomorphismLensesPost} \todo{more}.

@@

\begin{definition}
Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} is the category of optics with respect to this action: $\Prism = \Optic_\sqcup$.
\end{definition}

Just as prelenses correspond to a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms correspond to pairs of maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\Optic_\sqcup(S, A) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup A)l$

@@

\subsection{Preoptics}

@@

We can now show that profunctor preoptics are precisely preoptics in the ordinary sense.

@@

\begin{corollary}
A profunctor preoptic is determined by its component at $\Pastro_\M E_A$, and furthermore, this component is determined by its value on $\langle \id_A, \id_A, \id_A, \id_A \rangle \in (U \Pastro_\M E_A)(A, A)$.
\end{corollary}
\begin{proof}
This is the content of the first two isomorphisms above.

@@

\begin{proposition}
If $p : S \hto A$ is an optic then the associated natural transformation $(U-)(A,A) \to (U-)(S,S)$ is a lax monoidal. \todo{Rename the natural transformation $p$}
\end{proposition}
\begin{proof}
Let $\langle l, r \rangle$ be a representative for $p$ in $\Optic_\M(S,A)$ via the isomorphism of Proposition \ref{prop-profunctor-optics-are-optics}, so that
\begin{align*}
p_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
on each Tambara module $P$.

>>>>

\begin{remark}
In this section we will focus only on optics of the form $p : (S,S) \hto (A, A)$. We therefore abbreviate $\Optic_\M((S, S), (A, A))$ as $\Optic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.
\end{remark}

For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $outside : \Optic_\M(S, A) \to \C(S, S)$. By composition in the other direction we have a map $\C(S, M A) \times \C(M A, S) \to \C(M A, M A)$. Including this object $\C(M A, M A)$ into the coend $\int^{M \in \M} \C(M A, M A)$, we induce a map $inside : \Optic_\M(S, A) \to \int^{M \in \M} \C(M A, M A)$. 
\begin{definition}
An optic $p : S \hto A$ is \emph{lawful} if $outside(p) = \id_S$, and $inside(p) = \langle t A \rangle$ for some $t : M \to M$ in $\M$.
\end{definition}

\begin{proposition}
There is an subcategory $\Lawful_\C$ of $\Optic_\C$ given by objects of the form $(X, X)$ and lawful optics between them.
\end{proposition}
\begin{proof}
The identity optic $\iota(\id_X, \id_X)$ is clearly lawful.

Suppose we have two lawful optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$, and such that $lr = \phi S$ and $l'r' = \psi A$. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is also a lawful optic:
\begin{align*}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\intertext{and}
(Ml')l r (Mr') &= (Ml')(\phi S)(Mr') \\ 
&= (\phi NA) M(l')M(r') \\ 
&= (\phi NA) M(l'r') \\
&= (\phi NA) M(\psi A) \\
&= \phi(\psi A)
\end{align*}
by the naturality of the action of $\M$.
\end{proof}

@@

In view of this Proposition, whenever we posit a representative $\langle l, r \rangle$ of an lawful optic, we may assume that $lr = \phi A$ on the nose, for some $\phi : M \to M$.

@@

\begin{definition}
Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ is the category of lawful optics with respect to this action: $\Lens = \Lawful_\times$.
\end{definition}

As remarked in the introduction \todo{(todo: remark in the introduction)}, the optics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\Optic_\times(S, A) &= \int^{M \in \M} \C(S, M \times A) \times \C(M \times A, S) \\
&\cong \int^{M \in \M} \C(S, M) \times \C(S, A) \times \C(M \times A, S) && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A, S) && \text{(co-Yoneda)}
\end{align*}

@@

This has been written out many times before in the case when $lr$ is the identity \cite{IsomorphismLensesPost}, we note here that the weaker condition is sufficient.

@@

\begin{definition}
Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} is the category of lawful optics with respect to this action: $\Prism = \Lawful_\sqcup$.
\end{definition}

Just as (possibly unlawful) lenses correspond to a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$, (unlawful) prisms correspond to pairs of maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\Optic_\sqcup(S, A) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup A)l$

@@

\subsection{Optics}

@@

We can now show that profunctor optics are precisely optics in the ordinary sense.

@@

\begin{corollary}
A profunctor optic is determined by its component at $\Pastro_\M E_A$, and furthermore, this component is determined by its value on $\langle \id_A, \id_A, \id_A, \id_A \rangle \in (U \Pastro_\M E_A)(A, A)$.
\end{corollary}
\begin{proof}
This is the content of the first two isomorphisms above.

@@

\begin{proposition}
If $p : S \hto A$ is a lawful optic then the associated natural transformation $(U-)(A,A) \to (U-)(S,S)$ is a lax monoidal. \todo{Rename the natural transformation $p$}
\end{proposition}
\begin{proof}
Let $\langle l, r \rangle$ be a representative for $p$ in $\Optic_\M(S,A)$ via the isomorphism of Proposition \ref{prop-profunctor-optics-are-optics}, so that
\begin{align*}
p_P = (UP)(l,r) \alpha_{A,A,M}
\end{align*}
on each Tambara module $P$.
